### 2019-09-25 ch9_discussion/content.tex
update

A more significant and harder to solve problem is that the data formats produced by different methods varies greatly. This makes visualising and comparing multiple trajectories difficult, as different output types cannot be compared directly. The most commonly used and general is one where cells are positioned along a set of edges connecting milestones (Figure~\ref{fig:method_types}B). In practice, this format consists of two data structures: the milestone network specify transition between cell states, and the cell progressions specify how far along each cell has progressed along a transition (Figure~\ref{fig:method_types}C). Optionally, a cell can be in a region of "delayed commitment", that is, it is not known which state will end up in but a probability for ending up in either state can be assigned (Figure~\ref{fig:method_types}D).

\texttt{dynwrap} provides implementations for converting the output from different TI subtypes into the common format. Using this standardised format allows developing reusable software for visualising (\texttt{dynplot}) and comparing (\texttt{dyneval}) trajectories from different TI methods. All of these software packages are part of the \texttt{dyno} collection, soon to be available on CRAN \cite{dyno}.

@@

To solve this problem, benchmarking studies typically use synthetic data to complement real data. 
Several simulators have already been used to generate datasets to evaluate TI datasets, namely dyntoy \cite{saelens_comparisonsinglecelltrajectory_2019}}, dyngen \cite{dyngen}, splatter \cite{zappia_splattersimulationsinglecell_2017}, and PROSSTT \cite{papadopoulos_prossttprobabilisticsimulation_2018}.
Benefits of synthetic data are that they offer more control over the data characteristics, and that they can be generated in large quantities. This allows to evaluate performance of a method in function of a changing parameter (e.g. dataset size or noise levels), which provides information on how well the method will work on real datasets.

@@

By now, already hundreds of suitable real datasets exist (Figure~\ref{fig:datasets}). Downloading and pre-processing them requires a significant time investment. However, by processing the datasets once and storing them in a single repository allows them to be reused for multiple purposes. Readers are invited to reuse and extend the datasets used from our comparison of TI methods hosted on Zenodo\cite{cannoodt_singlecellomicsdatasets_2018}.

@@

%waarom benchmarks een laag aantal methoden hebben:
%* iedere methode heeft zijn eigen set van functies
%* methodes hebben vaak andere prior inputs (starting cell, time series, etc)
%* methodes zijn vaak in verschillende talen geschreven (R, python, matlab)
%* de output van methodes zijn vaak niet rechtstreeks vergelijkbaar, omdat er geen standaard definitie van wat trajectory inference is (zoals bv. wel het geval is bij clustering of differential expression)
%
%bovendien is trajectory inference een probleem met complexe output, waardoor geen kant-en-klare metrieken bestonden om de predicties mee te evalueren. ik zou nog eens door alle manuscripten die een evaluatie hebben gedaan, moeten gaan, maar ik denk dat alle of toch de meeste evaluaties tussen 2016 en 2018 enkel op basis van pseudotime zijn ─ ook voor bv. dpt, en dat er maar zelden evaluaties zijn die echt gaan kijken naar de branching
%
%
%kaart aan dat we in 2018 een benchmark preprint hebben gepost van W methoden op X datasets, en in 2019 een artikel hebben gepubliceerd van Y methoden op Z datasets. Ondanks dat we in deze manuscripten de grootste problemen (# datasets, # methodes, # metrieken, leg iets beter uit) aanpakken, zien we geen vermindering in het aantal nieuwe publicaties die geen self-benchmark bevatten.
%in deze commentary geven we nog een kort overzicht van:
%* standaard definitie van een trajectory inference probleem
%* waar kan je datasets vinden (dynbenchmark zenodo, de verschillende simulatoren, of download ze zelf van GEO)
%* hoe kan je methoden vergelijken (i.e. gebruik dynmethods)
%* wat zijn bestaande metrieken (toon overzicht van metrieken uit verschillende benchmarks en ook die van dyneval)
%in conclusie, met al deze informatie melden we dat in 2019 er dus geen enkele reden is om ons te gedragen als gelijk welk ander computationeel / data mining / machine learning veld waarbij er voor quantitatief bewijs wordt gevraagd dat je methode goed werkt als je een nieuwe methode wilt publiceren. we vragen vriendelijk aan peer-reviewers om kritisch te zijn over deze kwestie.

>>>>

A more significant and harder to solve problem is that the data formats produced by different methods varies greatly. This makes visualising and comparing multiple trajectories difficult, as different output types cannot be compared directly. The most commonly used and general is one where cells are positioned along a set of edges connecting milestones ("Regular TI", Figure~\ref{fig:method_types}A). 
By adding an extension to regular TI to allow for cells to be part of three or more cellular states, thereby a cell to delay its commitment toward a particular end state (Figure~\ref{fig:method_types}B).

By adding this extension, all TI subtypes can easily be converted into the common format. Implementations of these conversions can be found in \texttt{dynwrap}\cite{dyno}. Using this standardised format allows developing reusable software for visualising and comparing trajectories from different TI methods.

In practice, this format consists of two data structures: the milestone network specify transition between cell states, and the cell progressions specify how far along each cell has progressed along a transition (Figure~\ref{fig:method_types}C). In addition, regions of delayed commitment need to be specified, if any (Figure~\ref{fig:method_types}D).

@@

Several authors benchmark their method against synthetic data in order to evaluate their method \cite{haghverdi_diffusionmapshighdimensional_2015, street_slingshotcelllineage_2018,rizvi_singlecelltopologicalrnaseq_2017,welch_matchermanifoldalignment_2017,guo_hoplandsinglecellpseudotime_2017,boukouvalas_bgpbranchedgaussian_2017,gao_calistaclusteringlineage_2018,parra_reconstructingcomplexlineage_2018,chen_singlecelltrajectoriesreconstruction_2019,strauss_gpseudoclustdeconvolutionshared_2019}. These simulators are mostly custom made by the authors themselves, thereby increasing the level of bias in the benchmark. Instead, readers are encouraged to use of the readily available simulators, namely dyntoy \cite{saelens_comparisonsinglecelltrajectory_2019}, dyngen \cite{dyngen}, splatter \cite{zappia_splattersimulationsinglecell_2017}, and PROSSTT \cite{papadopoulos_prossttprobabilisticsimulation_2018}.

Benefits of synthetic data are that they offer more control over the data characteristics, and that they can be generated in large quantities. This allows to evaluate performance of a method in function of a changing parameter (e.g. dataset size or noise levels), which provides information on how well the method will work on real datasets.

@@

Several authors use mainly real datasets to evaluate their method, though only few use more than four datasets\cite{cannoodt_scorpiusimprovestrajectory_2016,liu_reconstructingcellcycle_2017,sharma_forksfindingorderings_2017,guo_hoplandsinglecellpseudotime_2017,teschendorff_singlecellentropyaccurate_2017,macnair_psupertimesupervisedpseudotime_2019}. 
By now, already hundreds of suitable real datasets exist (Figure~\ref{fig:datasets}). Downloading and pre-processing them requires a significant time investment, but by processing the datasets once and storing them in a single repository they can be reused for multiple purposes.

Readers are welcome to reuse (and extend) the 110 real and 229 synthetic datasets used in our comparison of TI methods. The datasets are hosted on Zenodo\cite{cannoodt_singlecellomicsdatasets_2018} and the scripts to process them on GitHub\footnote{\href{https://github.com/dynverse/dynbenchmark/tree/master/scripts/01-datasets}{github.com/dynverse/dynbenchmark/tree/master/scripts/01-datasets}}. Note that the ground truths of the datasets are represented using the common data structures format in the previous section.

@@

%in deze commentary geven we nog een kort overzicht van:
%* standaard definitie van een trajectory inference probleem
%* waar kan je datasets vinden (dynbenchmark zenodo, de verschillende simulatoren, of download ze zelf van GEO)
%* hoe kan je methoden vergelijken (i.e. gebruik dynmethods)
%* wat zijn bestaande metrieken (toon overzicht van metrieken uit verschillende benchmarks en ook die van dyneval)
%in conclusie, met al deze informatie melden we dat in 2019 er dus geen enkele reden is om ons te gedragen als gelijk welk ander computationeel / data mining / machine learning veld waarbij er voor quantitatief bewijs wordt gevraagd dat je methode goed werkt als je een nieuwe methode wilt publiceren. we vragen vriendelijk aan peer-reviewers om kritisch te zijn over deze kwestie.
